Clicks and Cliques

### 2018-12-28 papersoul.Rnw
include reviewer comments until p13, l40

# to be removed...
# library(reshape2)
# library(plyr)

@@

Data from ACS are comunity variable, computed using sampling weights by Census Bureau. 
When using data from Knight Foundation surveys, the sampling weights need to be taken into
account, if the data comes from only one year then the sampling weight is provided
in the data, when all three year survey data are used, sampling weights are divided by three 
so the total aggregate of each variable still can be consider an estimate of the comunity
total.

@@

The Knight survey data contains missing information. The proportion of missing values was computed for each observation (by variables) and for each variable (by cases). Table \ref{tab1} shows the quantiles of the proportion of missing values computed by case and computed by variable. Focusing on the proportions of missing values by variables, there is 19\% of missing record for the most complete response. This may be explained by the fact that 3 years of data was combined although some variables only appeared in 1 or 2 years. On the other hand, focusing on the proportion of missing values by casess, most of the variables show a reasonable proportion of missing values (the median is 12\%) while there are a few variables with really high proportions of missing cases.

@@

tab1 <- cbind(quantile(apply(miss[,orig], 1, mean) ),
            quantile(apply(miss[,reco], 1, mean)),
            quantile(apply(miss[,orig], 2, mean) ),
            quantile(apply(miss[,reco], 2, mean) ) )

@@

miss.c2 <- apply(miss[, miss.v < .25], 1, mean)
tab2 <- rbind(quantile(miss.v), quantile(miss.c) )
rownames(tab2) <- c('by Variable', 'by Case')
colnames(tab2) <- c('Min', 'Q25', 'Median', 'Q75', 'Max')
cap <- 'Quantiles of the proportion of missing values by case and variables.'
xt <- xtable(tab2, caption = cap, label = 'tab1', align = 'lccccc' )
print(xt, ,caption.placement = "top")

@@

Product plot is a very flexible framework \citep{wickham2011product}, which incorporate displays like bar charts, spine plots \citep{hummel1996linked} and mosaic plots \citep{hartigan1981mosaics,friendly1994mosaic}. This framework is designed for study relations among categorical variables. However, it is easily adapted when the variables are continuous simply by binning variables.

@@

A parallel coordinate plot \citep{inselberg1985plane} is used to compare the cluster means for the variables used in the analysis. Figure \ref{cluster.cen} shows the variables used in the horizontal axis and the variable mean of each cluster on the vertical axis. Group 1 are communities with older people, low education level and biggest home owner rate. Group 2 is formed by communities with large proportion of African-Americans and unequal income distribution. Bradenton and Palm Beach (entire group 3) are communities with older people, not working and not African-American. Group 4 is formed by four communities: Charlotte, Grand Forks, Long Beach and State College, all relatively younger people, with high education levels and non equitable income distribution.

@@

Using the original variables (with less than 25\% of missing values), k-means method are used to obtain the clusters of individuals. The variables were converted into a continuous scale. The exploratory search with k-means suggests that after four clusters the within variance reduction is small, so four clusters were used.

Figure \ref{para} is a parallel coordinate plot of the cluster means. The first to notice is that the cluster means are very similar among years.

@@

Cluster 1 shows the biggest mean in most of the variables. These people love their community, and feel it is a good place for different social groups. The only two variables in which this group does not present the largest means are the proportion of close friends and the proportion of family that live in the same community. On the other hand, cluster 2 seems to be the opposite to cluster 1, people in this group are not satisfied with the community where they live.

The only differences among means for clusters 3 and 4 are family and friends indices, i.e. the clique variables. Cluster 4 is characterized by the larger family and friends indices. Clusters 3 and 4 have quite similar mean values for the rest of the variables used to create the clusters.

@@

@
\caption{Relation between community attachment and k-means cluster solution. Cluster 1 shows the higest attachment proportion while cluster 2 the lowest. Attachment distribution is stable within each grupo amont years. \label{clus1}}
\end{figure}

@@

cors %>%  
  filter(vars %in% c("Prop.new.entry", "Average.years.rent.in.community", "Aver.years.own.in.community") &  type =='Attached' ) %>% 
  select(vars, vv) %>% rename(cor = vv) %>%
  bind_rows(data.frame(vars = 'stay', cor = rho)) %>% 
  mutate(Data =  c(rep('Census Bureau', 3), 'Knight-Gallup data'), 
         Description = c( 'Average years owning the current home', 
                          'Average years renting in current home', 
                          'Prop. of people not borned in the community', 
                          'Proportion of age in current comunity')) %>% 
  rename(Correlation = cor) %>% 
  select(Data, Description, Correlation) %>% 
  xtable( caption='Correlation with attachment', label='lifetab') %>% 
  print( caption.placement = 'top', include.rownames = FALSE)
@

@@

<<moving, dependson='data', echo=FALSE,results='asis', cache=FALSE>>=
soul %>% 
  mutate(Q5 = recode_factor(Q5, "(DK)" = "Non Response", 
    "(Refused)" = "Non Response",
    "Move outside of (answer from Sc) but stay in the state" = "Move from your neighborhood",
    "Move outside of this communitybut stay in the state" = "Move from your neighborhood",
    "Move to another city and state altogether" = "Move from your neighborhood",
    "Move to another neighborhood in (answer from Sc)" = "Move from your neighborhood", 
    "Move to another neighborhood in this community" = "Move from your neighborhood")) %>%  
  filter(!is.na(PROJWT)) %>% 
  mutate(pr = PROJWT/3)  %>% 
  svydesign(ids =~1, strata =~QSB, weights =~pr, data = .) %>% 
  svyby(~Q5,~CCEGRP2, design = ., svymean,keep.var = FALSE, na.rm = T) %>% 
  ftable() %>% data.frame() %>% mutate(Freq = Freq*100) %>% 
  mutate( Var2 = gsub('statistic.Q5', '', Var2) ) %>% 
  spread(CCEGRP2, Freq) %>%  
  xtable(caption = 'Willingness to move outside your neighborhood', label = 'move') %>% 
  print(caption.placement = 'top', include.rownames=FALSE)

@@

<<race, echo=FALSE, dependson=c( 'data' ), message=FALSE, fig.height=4>>=

@@

soul %>% select(race, CCEGRP2, CCE, QSB,WEIGHT, YEAR) %>% 
  filter(!is.na(WEIGHT), YEAR!=2010)  %>% 
  mutate(wt = WEIGHT/2) %>% 
  svydesign(ids=~1, strata=~QSB, weights=~wt, data=.) %>% 
  svyby(~race+CCEGRP2+CCE,~QSB, design = ., svymean,keep.var=FALSE, na.rm=T) %>%
  gather(race, prop, contains('race')) %>% 
  group_by(QSB) %>% 
  summarise(
    statistic.CCEGRP2Attached = first(statistic.CCEGRP2Attached),
    diver  = -sum( prop[prop > 0]*log(prop[prop>0]) )
  ) %>% 
  ggplot(aes(x = diver, y = logit(statistic.CCEGRP2Attached))) +
  geom_point(size = I(5)) + 
  geom_smooth(se = FALSE, method = 'lm', size = I(1)) +
  theme(axis.text.y = element_text( size = I(10)), aspect.ratio = 1 ) + 
  labs( y = 'Proportion of Attach (logit scale)', x = 'Race diversity index')

# b <- merge(census.data,tab[,c('QSB','diver','statistic.raceWhite') ], by='QSB') # why?
# rho.div <- with(tab, round(cor(diver, statistic.CCE),2)) # what is used for??

# wt <- with(subset(soul, !is.na(PROJWT) & YEAR!=2010), PROJWT/2 )
# des.pro <- svydesign(ids=~1, strata=~QSB, weights=~wt, data=subset(soul, !is.na(PROJWT) & YEAR!=2010))
# 
# tab <- svyby(~race+CCEGRP2+CCE,~QSB, des.pro, svymean,keep.var=FALSE, na.rm=T)
# tab$diver <- apply(tab[,grep('race',colnames(tab))], 1, function(x) -sum(x[x!=0]*log(x[x!=0])))
# b <- merge(census.data,tab[,c('QSB','diver','statistic.raceWhite') ], by='QSB')
# 
# rho.div <- with(tab, round(cor(diver, statistic.CCE),2))
# 
# ggplot(data = tab, aes(x = diver, y = logit(statistic.CCEGRP2Attached))) + geom_point(size = I(5)) + theme(axis.text.y = element_text( size = I(10)) ) + labs( y = 'Proportion of Attach (logit scale)', x = 'Race diversity index') + geom_smooth(se = FALSE, method = 'lm', size = I(1))

@@

Some variable can be identified as relatively most important for communities with high levels of attachment compare to communities with lower attachment levels. For instance, the columns corresponding to quality of healthcare (Health) or quality of college and universities (College) in the community show red boxes at the top of the plot.  On the other hand, variables relatively more important for the bottom communities (with lower levels of attachment) are related with economic conditions (economy) and the retrospective vision of the community (5.years.ago). Additionally there are other variables which its relative importance is not associated with attachment level, 5.years.ahead or Beauty presents red boxes for several communities with different levels of attachment.

@@

\bibliographystyle{spr-chicago}
\bibliography{papersoul}

@@

\section{Description of the variables used}

@@

\caption{Variables used in the paper}

>>>>

@@

Data from ACS are community variable, computed using sampling weights by Census Bureau. 
When using data from Knight Foundation surveys, the sampling weights need to be taken into
account, if the data comes from only one year then the sampling weight is provided
in the data, when all three year survey data are used, sampling weights are divided by three 
so the total aggregate of each variable still can be consider an estimate of the community
total.

@@

The Knight survey data contains missing information. The proportion of missing values was computed for each observation (by variables) and for each variable (by cases). Table \ref{tab1} shows the quantiles of the proportion of missing values computed by case and computed by variable. Focusing on the proportions of missing values for each observation (by variables), there is 19\% of missing record for the most complete response. This may be explained by the fact that 3 years of data was combined although some variables only appeared in 1 or 2 years. On the other hand, focusing on the proportion of missing values for each variable (by cases), most of the variables show a reasonable proportion of missing values (the median is 12\%) while there are a few variables with really high proportions of missing cases.

@@

@@

miss.c2 <- apply(miss[, miss.v < .25], 1, mean)
tab2 <- rbind(quantile(miss.v), quantile(miss.c) )
rownames(tab2) <- c('by cases', 'by variables')
colnames(tab2) <- c('Min', 'Q25', 'Median', 'Q75', 'Max')
cap <- 'Quantiles of the proportion of missing values by case and variables.'
xt <- xtable(tab2, caption = cap, label = 'tab1', align = 'lccccc' )
print(xt, ,caption.placement = "top")

@@

Product plot is a very flexible framework, which incorporate displays like bar charts, spine plots \citep{hummel1996linked} and mosaic plots \citep{hartigan1981mosaics,friendly1994mosaic}. This framework is designed for study relations among categorical variables. However, it is easily adapted when the variables are continuous simply by binning variables.

@@

A parallel coordinate plot \citep{inselberg1985plane} is used to compare the cluster means for the variables used in the analysis. Figure \ref{cluster.cen} shows the variables used in the horizontal axis and the variable mean of each cluster on the vertical axis. Group 1 are communities with older people, low education level and biggest home owner rate. Group 2 is formed by communities with large proportion of African-Americans and unequal income distribution. Bradenton and Palm Beach (entire group 3) are communities with older people, not working and not African-American. Group 4 is formed by four communities: Charlotte, Grand Forks, Long Beach, State College and Tallahassee, all relatively younger people, with high education levels and non equitable income distribution. Group 3 and 4 present opposite values in most analyzed variables. For example Age.Fem, Unemp and No.workers are positive and have the biggest values for group 3 among all the groups while for group 4 those variables have the smallest values among all groups. Groups 1, 2 and 3 are similar in Yr.1821, Bachelor , Gini and New.entry and are different in this variabels to group 4.

@@

Using the original variables (with less than 25\% of missing values), k-means clustering is used to obtain the clusters of individuals. The variables were converted into a continuous scale. The exploratory search with k-means suggests that after four clusters the within variance reduction is small, so four clusters were used.

Figure \ref{para} is a parallel coordinate plot of the cluster means. The first to notice is that the cluster means are very similar among years.
Cluster 1 shows the biggest mean in most of the variables. These people love their community, and feel it is a good place for different social groups. The only two variables in which this group does not present the largest means are the proportion of close friends and the proportion of family that live in the same community. On the other hand, cluster 2 seems to be the opposite to cluster 1, people in this group are not satisfied with the community where they live.
The only differences among means for clusters 3 and 4 are family and friends indices, i.e. the clique variables. Cluster 4 is characterized by the larger family and friends indices. Clusters 3 and 4 have quite similar mean values for the rest of the variables used to create the clusters.

@@

@@

@
\caption{Relation between community attachment and k-means cluster solution. Cluster 1 shows the highest attachment proportion while cluster 2 the lowest. Attachment distribution is stable within each group among years. \label{clus1}}
\end{figure}

@@

cors %>%  
  filter(vars %in% c("Prop.new.entry", "Average.years.rent.in.community", "Aver.years.own.in.community") &  type =='Attached' ) %>% 
  select(vars, vv) %>% rename(cor = vv) %>%
  bind_rows(data.frame(vars = 'stay', cor = rho)) %>% 
  mutate(Data =  c(rep('Census Bureau', 3), 'Knight-Gallup data'), 
         Description = c( 'Average years owning the current home', 
                          'Average years renting in current home', 
                          'Prop. of people not borned in the community', 
                          'Proportion of age in current comunity')) %>% 
  rename(Correlation = cor) %>% 
  select(Data, Description, Correlation) %>% 
  xtable( caption='Correlation with attachment.', label='lifetab') %>% 
  print( caption.placement = 'top', include.rownames = FALSE)
@

@@

<<moving, dependson='data', echo=FALSE,results='asis', cache=FALSE>>=
soul %>% 
  mutate(Q5 = recode_factor(Q5, "(DK)" = "Non Response", 
    "(Refused)" = "Non Response",
    "Move outside of (answer from Sc) but stay in the state" = "Move from your neighborhood",
    "Move outside of this communitybut stay in the state" = "Move from your neighborhood",
    "Move to another city and state altogether" = "Move from your neighborhood",
    "Move to another neighborhood in (answer from Sc)" = "Move from your neighborhood", 
    "Move to another neighborhood in this community" = "Move from your neighborhood")) %>%  
  filter(!is.na(PROJWT)) %>% 
  mutate(pr = PROJWT/3)  %>% 
  svydesign(ids =~1, strata =~QSB, weights =~pr, data = .) %>% 
  svyby(~Q5,~CCEGRP2, design = ., svymean,keep.var = FALSE, na.rm = T) %>% 
  ftable() %>% data.frame() %>% mutate(Freq = Freq*100) %>% 
  mutate( Var2 = gsub('statistic.Q5', '', Var2) ) %>% 
  spread(CCEGRP2, Freq) %>%  
  xtable(caption = 'Willingness to move outside your neighborhood.', label = 'move') %>% 
  print(caption.placement = 'top', include.rownames=FALSE)

@@

<<race, echo=FALSE, dependson=c( 'data' ), message=FALSE>>=

@@

soul %>% select(race, CCEGRP2, CCE, QSB,PROJWT, YEAR) %>% 
  filter(!is.na(PROJWT), YEAR!=2010)  %>% 
  mutate(wt = PROJWT/2) %>% 
  svydesign(ids=~1, strata=~QSB, weights=~wt, data=.) %>% 
  svyby(~race+CCEGRP2+CCE,~QSB, design = ., svymean,keep.var=FALSE, na.rm=T) %>%
  gather(race, prop, contains('race')) %>% 
  group_by(QSB) %>% 
  summarise(
    statistic.CCEGRP2Attached = first(statistic.CCEGRP2Attached),
    diver  = -sum( prop[prop > 0]*log(prop[prop>0]) )
  ) %>% 
  ggplot(aes(x = diver, y = logit(statistic.CCEGRP2Attached))) +
  geom_point(size = I(5)) + 
  geom_smooth(se = FALSE, method = 'lm', size = I(1)) +
  theme(axis.text.y = element_text( size = I(10)) ) + 
  labs( y = 'Proportion of Attach (logit scale)', x = 'Race diversity index')

@@

Some variable can be identified as relatively most important for communities with high levels of attachment compare to communities with lower attachment levels. For instance, the columns corresponding to quality of healthcare (Health) or quality of college and universities (College) in the community show red boxes at the top of the plot.  On the other hand, variables relatively more important for the bottom communities (with lower levels of attachment) are related with economic conditions (Economy) and the retrospective vision of the community (5.years.ago). Additionally there are other variables which its relative importance is not associated with attachment level, 5.years.ahead or Beauty presents red boxes for several communities with different levels of attachment.

@@

\bibliographystyle{spr-chicago}
\bibliography{papersoul}

@@

\section{Description of the variables created}

@@

\caption{Variables created in the paper.}
